---
title: "hw2_fb2731"
author: fan bu
output: github_document
date: "2026-10-07"
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(janitor)
```
# problem 1
```{r}
subway_raw =
  read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")
glimpse(subway_raw)
names(subway_raw)
subway_df =
  subway_raw |>
  clean_names()
names(subway_df)
subway_df =
  subway_df |>
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    starts_with("route"),
    entry,
    vending,
    entrance_type,
    ada
  )
glimpse(subway_df)
subway_df =
  subway_df |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
count(subway_df, entry)
dim(subway_df)

station_count =
  subway_df |>
  distinct(line, station_name) |>
  nrow()

station_count
count(subway_df, ada)
ada_station_count =
  subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()

ada_station_count
count(subway_df, vending)
no_vending_entry_prop =
  subway_df |>
  filter(vending == "NO") |>
  summarize(
    proportion = mean(entry, na.rm = TRUE)
  ) |>
  pull(proportion)

no_vending_entry_prop

subway_long =
  subway_df |>
  mutate(
    across(starts_with("route"), as.character)
  ) |>
  pivot_longer(
    cols = starts_with("route"),
    names_to = "route_number",
    values_to = "route_name"
  ) |>
  filter(!is.na(route_name))

a_train_station_count =
  subway_long |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

a_train_station_count

a_train_ada_count =
  subway_long |>
  filter(
    route_name == "A",
    ada == TRUE
  ) |>
  distinct(line, station_name) |>
  nrow()

a_train_ada_count

```
```{r}
# Problem 2

trash_file = "data/202610 Trash Wheel Collection Data.xlsx"

excel_sheets(trash_file)


mr_trash =
  read_excel(
    trash_file,
    sheet = 1,
    range = "A2:N1000"
  ) |>
  clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    month = as.character(month),
    year = as.integer(parse_number(as.character(year))),
    date = as.character(date),

    across(
      any_of(c(
        "weight_tons",
        "volume_cubic_yards",
        "plastic_bottles",
        "polystyrene",
        "cigarette_butts",
        "glass_bottles",
        "plastic_bags",
        "wrappers",
        "homes_powered"
      )),
      ~ parse_number(as.character(.x))
    ),

    sports_balls =
      as.integer(
        round(
          parse_number(as.character(sports_balls))
        )
      ),

    trash_wheel = "Mr. Trash Wheel"
  )



professor_trash =
  read_excel(
    trash_file,
    sheet = 2,
    range = "A2:N1000"
  ) |>
  clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    month = as.character(month),
    year = as.integer(parse_number(as.character(year))),
    date = as.character(date),

    across(
      any_of(c(
        "weight_tons",
        "volume_cubic_yards",
        "plastic_bottles",
        "polystyrene",
        "cigarette_butts",
        "glass_bottles",
        "plastic_bags",
        "wrappers",
        "sports_balls",
        "homes_powered"
      )),
      ~ parse_number(as.character(.x))
    ),

    trash_wheel = "Professor Trash Wheel"
  )



gwynnda_trash =
  read_excel(
    trash_file,
    sheet = 4,
    range = "A2:N1000"
  ) |>
  clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    month = as.character(month),
    year = as.integer(parse_number(as.character(year))),
    date = as.character(date),

    across(
      any_of(c(
        "weight_tons",
        "volume_cubic_yards",
        "plastic_bottles",
        "polystyrene",
        "cigarette_butts",
        "glass_bottles",
        "plastic_bags",
        "wrappers",
        "sports_balls",
        "homes_powered"
      )),
      ~ parse_number(as.character(.x))
    ),

    trash_wheel = "Gwynnda"
  )


trash_wheel_df =
  bind_rows(
    mr_trash,
    professor_trash,
    gwynnda_trash
  )

glimpse(trash_wheel_df)

trash_wheel_df |>
  count(trash_wheel)

nrow(trash_wheel_df)

professor_total_weight =
  trash_wheel_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  summarize(
    total_weight = sum(weight_tons, na.rm = TRUE)
  ) |>
  pull(total_weight)

professor_total_weight

gwynnda_june_2022 =
  trash_wheel_df |>
  filter(
    trash_wheel == "Gwynnda",
    year == 2022,
    month == "June"
  ) |>
  summarize(
    total_cigarette_butts =
      sum(cigarette_butts, na.rm = TRUE)
  ) |>
  pull(total_cigarette_butts)

gwynnda_june_2022

```


The combined Trash Wheel dataset contains `r nrow(trash_wheel_df)` observations. Each observation represents a dumpster collection from Mr. Trash Wheel, Professor Trash Wheel, or Gwynnda. Key variables include dumpster number, collection date, weight, volume, cigarette butts, plastic bottles, and Trash Wheel identifier.
Professor Trash Wheel collected a total of `r round(professor_total_weight, 2)` tons of trash. Gwynnda collected `r gwynnda_june_2022` cigarette butts in June 2022.

```{r}
# Problem 3

baseline_df =
  read_csv(
    "data/MCI_baseline.csv",
    skip = 1,
    na = c("", ".", "NA")
  ) |>
  clean_names()

amyloid_df =
  read_csv(
    "data/mci_amyloid.csv",
    skip = 1,
    na = c("", ".", "NA")
  ) |>
  clean_names()
baseline_df =
  baseline_df |>
  mutate(
    id = as.integer(id),
    current_age = as.numeric(current_age),
    education = as.numeric(education),
    age_at_onset = as.numeric(age_at_onset),

    sex = case_match(
      sex,
      0 ~ "Female",
      1 ~ "Male"
    ),

    apoe4 = case_match(
      apoe4,
      0 ~ "Non-carrier",
      1 ~ "Carrier"
    )
  ) |>
  filter(
    is.na(age_at_onset) |
      age_at_onset > current_age
  )

glimpse(baseline_df)

count(baseline_df, sex)
count(baseline_df, apoe4)

n_participants =
  nrow(baseline_df)

n_participants

n_mci =
  baseline_df |>
  filter(!is.na(age_at_onset)) |>
  nrow()

n_mci

mean_baseline_age =
  baseline_df |>
  summarize(
    mean_age = mean(current_age, na.rm = TRUE)
  ) |>
  pull(mean_age)

mean_baseline_age

female_apoe4_prop =
  baseline_df |>
  filter(sex == "Female") |>
  summarize(
    proportion = mean(apoe4 == "Carrier", na.rm = TRUE)
  ) |>
  pull(proportion)

female_apoe4_prop

female_apoe4_prop * 100

amyloid_long =
  amyloid_df |>
  mutate(
    across(
      c(baseline, time_2, time_4, time_6, time_8),
      as.numeric
    )
  ) |>
  pivot_longer(
    cols = c(baseline, time_2, time_4, time_6, time_8),
    names_to = "visit",
    values_to = "amyloid_42_40"
  ) |>
  mutate(
    time = case_match(
      visit,
      "baseline" ~ 0,
      "time_2" ~ 2,
      "time_4" ~ 4,
      "time_6" ~ 6,
      "time_8" ~ 8
    )
  ) |>
  select(
    study_id,
    time,
    amyloid_42_40
  )

glimpse(amyloid_long)

baseline_only =
  baseline_df |>
  anti_join(
    amyloid_long |>
      distinct(study_id),
    by = c("id" = "study_id")
  )

baseline_only
nrow(baseline_only)

amyloid_only =
  amyloid_long |>
  distinct(study_id) |>
  anti_join(
    baseline_df,
    by = c("study_id" = "id")
  )

amyloid_only
nrow(amyloid_only)
mci_combined =
  amyloid_long |>
  inner_join(
    baseline_df,
    by = c("study_id" = "id")
  )

glimpse(mci_combined)

nrow(mci_combined)

n_distinct(mci_combined$study_id)

write_csv(
  mci_combined,
  "data/mci_combined.csv"
)


```

The baseline demographic dataset was imported after skipping the first
descriptive header row. Sex and APOE4 carrier status were recoded as
categorical variables, and participants who had developed MCI at or before
baseline were excluded. The cleaned baseline dataset contains
`r n_participants` eligible participants, of whom `r n_mci` developed MCI
during follow-up. The mean baseline age was
`r round(mean_baseline_age, 2)` years. Among female participants,
`r round(100 * female_apoe4_prop, 1)`% were APOE4 carriers.

The amyloid dataset contains amyloid 42/40 ratio measurements at baseline
and at 2, 4, 6, and 8 years of follow-up. I reshaped the data from wide to
long format so that each row represents one participant at one measurement
time.

There were `r nrow(baseline_only)` participants appearing only in the
baseline dataset and `r nrow(amyloid_only)` participants appearing only in
the amyloid dataset. The final dataset was created using an inner join so
that only participants appearing in both datasets were retained.

The final combined dataset contains `r nrow(mci_combined)` longitudinal
observations from `r n_distinct(mci_combined$study_id)` unique participants.